
# Working with data

## What are data?

The first important point about data is that data are - meaning that the word "data" is plural (though some people disagree with me on this).  You might also wonder how to pronounce "data" -- I say "day-tah" but I know many people who say "dah-tah" and I have been able to remain friends with them in spite of this. Now if I heard them say "data is" then that would be bigger issue...

### Qualitative data
Data are composed of *variables*, where a variable reflects a unique measurement or quantity.  Some variables are *qualitative*, meaning that they describe a quality rather than a numeric quantity.  For example, in my stats course I generally give an introductory survey, both to obtain data to use in class and to learn more about the students. One of the questions that I ask is "What is your favorite food?", to which some of the answers have been: blueberries, chocolate, tamales, pasta, pizza, and mango.  Those data are not intrinsically numerical; we could assign numbers to each one (1=blueberries, 2=chocolate, etc), but we would just be using the numbers as labels rather than as real numbers; for example, it wouldn't make sense to add the numbers together in this case.  However, we will often code qualitative data using numbers in order to make them easier to work with, as you will see later.

### Quantitative data
More commonly in statistics we will work with *quantitative* data, meaning data that are numerical.  For example, here Table \@ref(tab:WhyTakingClass) showing the results from another question that I ask in my introductory, which is "Why are you taking this class?"

```{r WhyTakingClass,echo=FALSE}

knitr::kable(
  head(mtcars[, 1:8], 10), booktabs = TRUE,
  caption = 'REPLACE WITH CORRECT TABLE.'
)
```

Note that the students' answers were qualitative, but we generated a quantitative summary of them by counting how many students gave each response.  

#### Types of numbers
There are several different types of numbers that we work with in statistics.  It's important to understand these differences, in part because programming languages like R often distinguish between them.

**Binary numbers**. The simplest are binary numbers -- that is, zero and one.  We will often use binary numbers to represent whether something is true or false, or present or absent.  For example, I might ask 10 people if they have ever experienced a migraine headache.  If their answers were:

```{r}
everHadMigraine = c('Yes','No','Yes','No','No','No','Yes','No','No','No')
everHadMigraine
```

I could instead recode these into truth values:

```{r}
everHadMigraineTF = everHadMigraine=='Yes'
everHadMigraineTF
```

R treats truth values and binary numbers eqivalently:

```{r}
TRUE==1
FALSE==0
TRUE==0
```

And we can also turn our list of truth values into integers explicitly:

```{r}
everHadMigraineBinary=as.integer(everHadMigraineTF)
everHadMigraineBinary
```

When we get to probability theory we will see a way in which this kind of representation can be very useful.
 
**Integers**.  Integers are whole numbers with no fractional or decimal part. We most commonly encounter integers when we count things, but they also often occur in psychological measurement.  For example, in my introductory survey I administer a set of questions about attitudes towards statistics (such as ""), on which the students respond with a number between 1 ("") and 7 ("").  

**Real numbers**.  Most commonly in statistics we work 

